Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-06-04 WSSSPE2_report/WSSSPE2_review.tex
Update WSSSPE2_review.tex
Additions and updates to Exploring Sustainability section.

Each author was invited to outline the key action from their paper as a potential discussion point for the group; in
their absence, the group facilitators outlined the actionable points from papers where the authors were not present.
The key actions from the six papers included recommendations for improving practice in sustainable software
engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service
as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributions
to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in
scalability, reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software architectures as a mechanism
for architectural-level reasoning about sustainability~\cite{wssspe2_venters}. The group took a position to view
sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of
reusable software~\cite{XXX}. \katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards
developing sustainable software in academic environments. It was suggested that a new requirement driven by the agencies
funding research projects where software was an intrinsic part of enabling the research program would be an additional
element within the grant proposal to provide a sustainability plan for sustaining the software. This type of initiative would
provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime
of the project.

>>>>

Each author was invited to outline the key action from their paper as a potential discussion point for the group; in
their absence, the group facilitators outlined the actionable points from papers where the authors were not present.
The key actions from the six papers included recommendations for improving practice in sustainable software
engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service
as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributionsto open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in
scalability, reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software architectures as a mechanism
for architectural-level reasoning about sustainability~\cite{wssspe2_venters}. The group took a position to view
sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of
reusable software~\cite{XXX}. \katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards developing sustainable software in academic environments. It was suggested that a new requirement driven by the agencies funding research projects where software was an intrinsic part of enabling the research program would be an additional element within the grant proposal to provide a sustainability plan for sustaining the software. This type of initiative would provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime of the project.

The discussion then focused on a need for a common language and a definition on the concept of sustainability that moved beyond the current fuzzy definitions where time was the simple measurement of sustainment. It was suggested that there was a need to identify tangible actions that underpin sustainability that developers could incorporate into the development stream of their software. This prompted a debate regarding whether sustainability should be considered as a non-functional requirement or software quality as defined within ISO/IEC 25010~\cite{}. The focus of the discussion was based on the paper by Venters et. al., ~\cite{wssspe2_venters}, which suggested that sustainability is a first-class, composite, non-functional requirement composed of a number of sub-characteristics. It was generally agreed that maintainability and extensibility were key qualities underlying sustainability. In addition, the group also discussed what other non-functional requirements would contribute to the development of sustainable software e.g. reusability and scalability. However, it was recognized that there was a need to identify appropriate metrics and measures.

The group also discussed whether the concept of sustainability was a barrier in itself to achieving sustainable software. It was suggested that "the first rule of software sustainability is, do not to talk about software sustainability". Instead there should be a focus on best software engineering practice. In order to play devils advocate, it was suggested that if the focus on the concept of sustainability was ignored, what current software engineering practices and principles could be utilized by software developers and domain scientists to achieve sustainability. This raised the question of why existing software engineering knowledge, such as that contained in SWEBOK~\cite{} was largely ignored and to what extent the environment has a strong influence on practice within the scientific and engineering community. As a result, there is a need to identify best practice and reach a consensus beyond the theories that underpinned the discipline of software engineering. Similarly, how could we translate or distill some of the key building blocks that underpin software engineering practice.

A final key point of the discussion was the role that software design and patterns play before committing to writing a line of code. It was suggested that modeling must play a major role in attaining sustainable software. The point was raised that design involves making decisions and a mechanism would be necessary for capturing these decision points. This introduced the idea of software provenance that moved beyond commits in software repositories to how to capture and maintain relationships between sources and design decisions to prevent knowledge vaporization. Whether this could be achieved through software architectures was a open research challenge.
